---
title: "Assignment 7"
author: "Cassie Boylan"
date: "10/10/2021"
output: html_document
---

```{r setup, include=FALSE, warning=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning=FALSE)

```

```{r message=FALSE}
library("XML")
library("methods")
library("rjson")
library("httr")
library("htmltab")
library("RCurl")
```

## Introduction
The purpose of this assignment is to familiarize ourselves with the structures of json, extended markup language and html and how each stores objects and attributes.  We generated a simple dataset of books with attributes such as title, author and publish date and created a separate file for each language to represent these books and attributes contained.  These files were then read into r utilizing specific packages for this purpose and transformed into tidy dataframes for further use.

```{r json}
#read json file into R
books_json <- fromJSON(file="https://raw.githubusercontent.com/cassie-boylan/DATA-607/main/books_json_version.json")

#Convert json output to data frame
books_json_data_frame <- as.data.frame(books_json)
print(books_json_data_frame)
```

```{r xml}

#Convert xml output to data frame
books_xml_dataframe <- xmlToDataFrame("booksv2.xml")
print(books_xml_dataframe)
```

```{r html}
books_html <- htmltab("https://raw.githubusercontent.com/cassie-boylan/DATA-607/main/books.html", which =1)

print(books_html)

```

