---
title: "Project 2"
author: "Cassie Boylan"
date: "10/3/2021"
output: 
    prettydoc::html_pretty:
    theme: cayman
---

```{r setup, include=FALSE, warning=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(ggplot2)
library(dplyr)
library(stringr)
library(readr)
library(data.table)
library(visdat)
library(ggrepel)
```


# Dataset 1 - Global TB rates

This dataset represents the tuberculosis cases reported in the past 30 years by country and demographic.  The original dataset has multi-variate columns and is missing a substantial amount of data.

I will be tidying this dataset so that each row represents a single observation (i.e. the cases of tb for a particular country/year/gender, & age range).  I will be identifying if there are any patterns in the data that is missing and dropping or imputing NA values accordingly.  I will be looking up country name to replace the non-intuitive iso country code in the original dataset.

I will then perform some EDA and summary statistics on the tidied dataset.

## Importing the Data
```{r}
NAMES_tb <- read.table("https://raw.githubusercontent.com/rodrigomf5/Tidydata/master/tb.csv", nrow = 1, stringsAsFactors = FALSE, sep = ",")

DATA_tb <- read.table("https://raw.githubusercontent.com/rodrigomf5/Tidydata/master/tb.csv", skip = 1, stringsAsFactors = FALSE, sep = ",")

tb <- DATA_tb[, 1:length(NAMES_tb)]
names(tb) <- NAMES_tb
```
[credit for structure]:
(https://stackoverflow.com/questions/38311808/more-columns-than-column-names/38311962)
```{r}
lookup_country <- read_csv("https://pkgstore.datahub.io/core/country-list/data_csv/data/d7c9d7cfb42cb69f4422dec222dbbaa8/data_csv.csv", show_col_types = FALSE)
```

```{r}
glimpse(tb)
```
## Reshaping the Dataset

I am collapsing the columns that are storing multiple variables per column (gender, age range, and tb_cases) into 2 columns - demographic and case count.  I am pivoting on iso2, year, and new_sp columns.

I chose to keep new_sp column as this column captures the total of all tb cases reported per country per year.  I am renaming this column total_cases for better accuracy.
```{r}
tb_clean <- tb %>%
  pivot_longer(
    cols = -c(iso2,year,new_sp),
    names_to = "demo", 
    values_to = "cases") %>%   
  rename("total_cases" = new_sp) 
```

Since the disparate variables of the demo (demographic) column are not separated by any non-alphanumeric characters, I am using str_extract() inside a mutate to pull out variables gender and age range instead of a separate() verb.

Both new variables will be factor variables in that they have a fixed select number of values they could be, so I am also casting both variables to factor inside my mutate.

```{r}
tb_clean <- tb_clean %>%
  mutate(
    gender = as.factor(str_extract(demo, "f|m")),
    age_range =                              (str_extract(demo,"[0-9]{2,4}"))
    ) %>%
  replace_na(list(age_range = "unknown")) %>%
  select(-demo)
```
## Handling Missing Data

It is noticeable that there are many NA values in this dataset.
```{r}
sum(is.na(tb_clean))

```

```{r}
vis_miss(tb_clean)
```

Looking at the data, I suspect that the NA values are more often happening in earlier years where data may have been more difficult to ascertain.  I am going to arrange by year and then call vis_miss.  Should my suspicion be correct, the missing values should all show as sequential (ie. data is missing at random but not completely at random).
```{r}
tb_clean_by_year <- tb_clean %>%
  arrange(year)
vis_miss(tb_clean_by_year)
```

It looks like my hunch was correct.  To see at around what year did frequency of NA values drop off, I will filter and count only the observations with NA in the cases column and then plot.

```{r}
tb_missing <- tb_clean %>%
  filter(is.na(cases)) %>%
  count(year, cases, name = "freq")

ggplot(tb_missing, aes(x=year, y=freq)) + geom_col()
```

It appears that only around 1995 did data begin to be collected in a comprehensive manner.  It could still be helpful to retain incidence rates from the 80s for certain countries where this data was collected successfully.  However, I am looking to apply summary statistics across countries, and the quality/accuracy of this analysis will significantly decrease with so many missing values; therefore I am going to drop all observational records from 1980-1995.
```{r}
tb_clean$year <- as.numeric(tb_clean$year)

exclude <- c(1980:1995)

tb_clean<- tb_clean %>%
  filter(!(year %in% exclude))
```

At this point, we have the choice to either assume that where values are NA in the case column, no tb cases were reported *or* no data was collected at all.  We can discern from the dataset that there do exist records of cases reported at 0 indicating that NA in this dataset is not necessarily a proxy for 0 cases. 
```{r}
tb_clean %>% filter(cases == 0) %>% count(cases)
tb_clean %>% filter(is.na(cases)) %>% count(cases)
```

Where total cases reported are also NA, records containing NA values in cases variable will be dropped.  Where a total case count for country/year has been reported, cases will be imputed as 0.
```{r}
tb_clean <- tb_clean %>% filter(!is.na(cases) & !is.na(total_cases))

tb_clean <- tb_clean %>% replace_na(list(cases = 0L))

```
It was noted in our plot of missing values that there are a number of observations where the country code is missing.  After dropping by year and dropping by cases = NA, there is only 184 records without an iso code.  Reviewing this extracted data, it appears these values are not missing at random but are of a country code not included.  We will need to drop all records missing the iso code as imputation does not appear possible.
```{r}
sum(is.na(tb_clean$iso2))
missing_country <- tb_clean %>% filter(is.na(iso2))
missing_country %>% group_by(year) %>% tally()


```

## Replacing Variable Values via Lookup

By importing the lookup table for country and iso codes, we can lookup and add the country name for each iso code listed via join().  This will make for easier interpretation under which country each tb case record is reporting.
```{r}
tb_clean <- lookup_country %>% inner_join(tb_clean, by= c("Code"="iso2"))

tb_clean <- tb_clean %>% rename("country"= Name) %>% select(-Code)
```
## Exploratory Data Analysis on Age Ranges

If we list out the possible age ranges, we see that there is an age range value for unknown and age ranges for 0-4, 5-14, and 0-14 years of age.  Lets do some digging on the cases reported in these unexpected age ranges.
```{r}
tb_clean$age_range <- as.factor(tb_clean$age_range)

unique(tb_clean$age_range)
```
### TB Cases Reported where Age Range Unknown
How many cases have been reported where the gender but not the age range was unknown?  
Where/when were these cases highest?
```{r}
tb_clean %>% filter(age_range =="unknown" & cases > 0) %>% arrange(desc(cases))
```
It appears there are 53 cases (within the subset timeframe_ in which tb cases were reported absent of age range.  The largest numbers were reported in Nepal in 2008.  It also appears that Italy has in multiple recent years reported a small number of cases absent of age range information.

### TB Cases Reported in Children
How many cases of TB have been reported in the age range 0-4 years?  It appears that cases involving children 0-4 have only been reported in the last years of this dataset: 2005-2008.  The highest numbers recorded in any one year / country / gender were ~600 in South Africa 2006.  Within the same period, 2006-2008, tb cases were reported at their highest among children 5-14 in the countries of South Africa and Indonesia.
However, on the bright side, tb cases among children represent less than 2% of all reported cases in any year.
```{r }
zero_to_four <- tb_clean %>% filter(age_range== "04" & cases > 0) %>%
  mutate(percentage = scales::percent(cases/total_cases))

zero_to_four %>% arrange(desc(cases))
summary(zero_to_four[c("cases","gender")])
zero_to_four %>% group_by(country) %>% tally() %>% arrange(desc(n))
```

```{r}
five_to_fourteen <- tb_clean %>% filter(age_range=="514" & year %in% (2004:max(year)) & cases > 0) %>%
    mutate(percentage = scales::percent(cases/total_cases))

five_to_fourteen %>% arrange(desc(cases))
summary(five_to_fourteen[c("cases","gender")])
five_to_fourteen %>% group_by(country) %>% tally() %>% arrange(desc(n))
```

```{r}
zero_to_four %>% ggplot(aes(x=year,y=cases,color=gender)) + geom_point()

five_to_fourteen %>% ggplot(aes(x=year,y=cases,color=gender)) + geom_point()
```

## Calculating Total Cases

### Calculate total cases by country and year by summarizing cases reported within each demographic.  
Interesting enough, the manual summary of cases within each demographic by country/year is *not* equal to the total number of tb cases reported by country/year within the original dataset.  Comparing the difference, India appears to be most off in total cases vs cases totalled across demographics.  China, Ukraine, and Malaysia all appear to have cases of underreporting total tb cases vs cases totalled across demographics.
```{r}
summaries <- tb_clean %>% group_by(country, year) %>% summarize(country, year, total_cases, total = sum(cases), difference = total_cases - total) %>% unique()

summaries %>% arrange(desc(difference)) %>% head(10)
summaries %>% arrange(difference) %>% head(10)
```
## Summarizing Cases by Percentage of Yearly Total

It appears that reported tb case count has increased substantially in recent years.  It would be interesting to know if this is due to an actual increase in cases or the result of more accurate & complete reporting.
```{r}
by_year <- tb_clean %>%
  group_by(year) %>%
  mutate(yearly_total = sum(cases)) 

by_year %>%
  ggplot(aes(x=year,y=yearly_total)) + geom_line()

group_by_total <- by_year %>% 
  group_by(year,country) %>%
  summarize(
    percentage_of_all = round(total_cases/yearly_total,2)) %>%
  arrange(desc(percentage_of_all)) %>% 
  unique() %>%
  head(10)

group_by_total
```
It appears that China and India saw rapid increases in tb case reporting in the first decade of the 21st century. China and India both have had very high proportions of the yearly global tb case load in the past 2 decades.
```{r}

top_c <- by_year %>% filter(country %in% c("China","India","Indonesia"))

top_c %>% ggplot(aes(x=year,y=total_cases)) + geom_line(aes(color=country)) +scale_color_brewer(palette="Set1")
```

## Dataset 2 - Titanic
I am importing the popular dataset of survival rates among passenger demographics of the HMS Titanic sinking in 1910.

```{r}
NAMES <- read.table("https://raw.githubusercontent.com/cassie-boylan/DATA-607-Project-2/main/titanic.csv", nrow = 1, stringsAsFactors = FALSE, sep = ",")

DATA <- read.table("https://raw.githubusercontent.com/cassie-boylan/DATA-607-Project-2/main/titanic.csv", skip = 1, stringsAsFactors = FALSE, sep = ",", na.strings=c("", NA))

titanic <- DATA[, 1:12]
names(titanic) <- NAMES
```

## Extracting Multi-Variables
```{r}
titanic_clean <- titanic %>%
  separate(Name, into=c("Last.Name", "First.Name"), sep=",") %>%
  mutate(Age = round(Age,0))

names(titanic_clean)[8] <- "Sibling.Spouse.Count"
names(titanic_clean)[9] <- "Parent.Child.Count"

```

## Handling Missing Data

There are 177 observations missing data. Calling a plot of the dataset utilizing vis_dat,  I can see that all the NA values are within the variable Age and Cabin.

I can also see that all the dummy variables of this dataset are set as numeric. They should be represented as factors.
```{r}
sum(is.na(titanic_clean))
vis_dat(titanic_clean)
```

Since this age is a numeric value, imputing with the mean value seems reasonable.  I ran a quick five num and confirmed from IQR and confirmed that mean is similar to median and can be considered a fair measure of center.
```{r}
mean_age <- round(mean(titanic_clean$Age, na.rm=TRUE),0)

fivenum(titanic_clean$Age, na.rm=TRUE)

titanic_clean <- titanic_clean %>%
  mutate(Age = replace_na(Age, mean_age))
```

A quick look through the Cabin values does show that this variable does not carry any real meaning so I am dropping column.
```{r}
titanic_clean <- titanic_clean %>%
  select(-Cabin)
```

## How many of each gender & socio-economic class survived the Titanic crash?  

It appears that if you were a female of first or second class, your odds of survival look pretty good. Women of first and second class had a survival rate of 97% and 92%
Only 50% of woman of the third class survived, which was still better than men of the first class at 37%.  
Most saddening is the survival rate of men of the third class at 15%
```{r message=FALSE}
titanic_sum <- titanic_clean %>%
  group_by(Pclass, Sex) %>% 
  summarize(survived=round(sum(Survived)/n(),2),
            total_passengers = n(),
            dead_passengers = total_passengers - round((survived * total_passengers),0))

titanic_sum
```



# Datset 3 - Religion vs Income
This dataset tabulates survey responses of self-reported religious affiliation and yearly income.

## Importing Data
```{r}
religion_income <- read_csv("https://raw.githubusercontent.com/rodrigomf5/Tidydata/master/relinc.csv", show_col_types = FALSE)
```
## Reshaping Data
moving religion_income dataset long to wide
```{r}
religion_income <- religion_income %>%
  gather(key="income", value="frequency", "<10k":"refused", -religion) %>%
  filter(religion != "refused") %>%
  arrange(desc(frequency))

religion_income
```

```{r}
unique(religion_income$religion)
```
## Recoding Variables for Cleaner Analysis
```{r}
religion_income <- religion_income %>%
  mutate(income_level = case_when(
    income %in% c("75-100k","100-150k", ">150k") ~ "Wealthy", 
    income %in% c("50-75k","40-50k","30-40k") ~"Middle Class", 
    income %in% c("20-30k","10-20k", "<10k")~ "Blue Collar",
    TRUE ~ "unknown"))
```

```{r}
by_class <- religion_income %>%
  group_by(income_level) %>%
  summarise(ppl_total = sum(frequency)) %>%
  arrange(desc(ppl_total))
```

```{r}
by_religion <- religion_income %>%
  group_by(religion) %>%
  summarise(ppl_total = sum(frequency)) %>%
  arrange(desc(ppl_total))
```

```{r}
religion_income2 <- religion_income %>%
  group_by(income_level, religion) %>%
  mutate(total = sum(frequency)) %>%
  select(religion, income_level, total) %>%
  arrange(religion, income_level) %>%
  unique()
```

```{r}
religion_income_wide <- religion_income2 %>% pivot_wider(names_from = religion, values_from = total)
```

```{r}
rel_inc_distr <- religion_income2 %>% group_by(religion) %>% mutate(sum = sum(total),                          percent = scales::percent(total/sum))

```
## Subsetting Data to DrillDown
```{r}
religion_income_ca <- rel_inc_distr %>%
  filter(religion == "Catholic") 

religion_income_ca %>% select(income_level, percent)
```

```{r}
religion_skeptics <- rel_inc_distr %>% filter(religion %in% c("Atheist","Agnostic","Jewish"))
```

```{r}
religion_protestant <- rel_inc_distr %>% filter(grepl("Prot",religion))

```
## Plotting Subsets
```{r}
ggplot(religion_skeptics, 
        aes(fill=income_level, x=religion, y=total)) + geom_bar(position="stack",stat="identity") + scale_color_brewer(palette="Pastel1")

ggplot(religion_protestant, 
        aes(fill=income_level, x=religion, y=total)) + geom_bar(position="stack",stat="identity") + scale_color_brewer(palette="Pastel1")
```


